Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-30 paper.tex
Pass through discussion

Nonetheless, there are numerous datasets that exist today
that would likely reap significant benefits from being deployed 
in a cloud-native fashion using Zarr. Object 
stores typically allow for individual objects (chunks, in
Zarr) to be associated with ``tags'', which can then be 
used to associate storage class, user access control
and encryption keys on a chunk-by-chunk basis. 
Aside from the performance benefits
we have focused on here provided by Zarr, the ability
to (for example) use high-performance storage for commonly
used data such as the variant position and 
more cost-effective storage classes
for infrequently used bulk QC data should provide 
significant operational benefits.
Granular access controls would similarly allow non-identifiable
variant-level data to be shared relatively freely,
with genotype and other data more tightly controlled
as required.
Even finer granularity is possible if samples are grouped according
to access level within chunks (i.e, by padding partially filled 
chunks and using an appropriate sample mask). 
Delegating access control to the cloud provider in this way
and providing client applications direct access to
the data in an efficiently analysable form, 
largely negates the 
need for web API access~\cite{kelleher2019htsget} 
and cryptographic container formats~\citep{senf2021crypt4gh}
in this setting.

The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
provided here are a necessary starting point for such cloud-native
biobank repositories, but significant investment and development
would be needed 
to provide a viable alternative to standard bioinformatics workflows. 
A few key pieces of relatively straightforward infrastructure, 
however, would provide a starting point  and be
sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) on a VCF Zarr dataset 
would speed up some common queries by orders of magnitude,
and provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.

The second key piece of infrastructure required at this large 
scale concerns distributed computation. Even with more efficient
columnar data representations, computations that required examining 
large volumes of per-call data must be distributed to complete 
in a reasonable amount of time. [Dask, Xarray, sgkit?]

% \item Vision. An ecosystem of tools that interoperate
% based on an open, cloud-based format where distribution is naturally
% managed by the stored chunks. Tools can be run on different datasets
% with minimal tuning, not by writing a full layer of orchestration.
% Tools read the data they need, not the entire dataset. Users read
% directly from a single canonical dataset, and do not maintain
% their own filtered copies. We suggest that the Zarr VCF approach
% we have provided would be a good starting point for designing
% such a compute platform.

% \end{itemize}

The Python data science ecosystem, 
built on the foundations of efficient array-oriented 
computing~\citep{harris2020array},
has a rich suite of powerful tools~\cite{
mckinney2010data,
rocklin2015dask,
kluyver2016jupyter,
hoyer2017xarray,
virtanen2020scipy} and is gaining increasing
popularity in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Zarr has the potential to act as a unified, cloud-native 
efficient storage layer, freeing the next generation of 
bioinformatics tools
from the traditional burdens of text parsing and format
conversion.

>>>>

Nonetheless, there are numerous datasets that exist today
that would likely reap significant benefits from being deployed 
in a cloud-native fashion using Zarr. Object 
stores typically allow for individual objects (chunks, in
Zarr) to be associated with ``tags'', which can then be 
used to associate storage class, user access control
and encryption keys. 
Aside from the performance benefits
we have focused on here provided by Zarr, the ability
to (for example) use high-performance storage for commonly
used data such as the variant position and 
more cost-effective storage classes
for infrequently used bulk QC data should provide 
significant operational benefits.
Granular access controls would similarly allow non-identifiable
variant-level data to be shared relatively freely,
with genotype and other data more tightly controlled
as required.
Even finer granularity is possible if samples are grouped according
to access level within chunks (i.e, by padding partially filled 
chunks and using an appropriate sample mask). 
Delegating access control to the cloud provider in this way
and providing client applications direct access to
the data in an efficiently analysable form, 
largely negates the 
need for web API access~\cite{kelleher2019htsget} 
and cryptographic container formats~\citep{senf2021crypt4gh}
in this setting.

The draft VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
provided here are a necessary starting point for such cloud-native
biobank repositories and open up many possibilities,
but significant investment and development would be needed 
to provide a viable alternative to standard bioinformatics workflows. 
Two initial directions for development, however, may provide 
a sufficient starting point to both
greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England
and motivate further development.
The first direction is to provide compatibility with existing 
workflows via a ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) on a VCF Zarr dataset.
Such a tool would speed up some common queries by orders 
of magnitude, and reduce the need for user orchestration of 
operations among manually split VCF chunks (large VCF datasets
are typically split into hundreds of files; see the Genomics
England case study). Datasets could then be hosted in cloud
object stores, while still presenting file-like semantics
for existing workflows. This could provide an evolutionary path,
allowing established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.

The second direction is then to develop new methods 
working directly with the Zarr data, and can leverage 
a rich ecosystem of powerful and efficient tools in 
multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
harris2020array,
mckinney2010data,
kluyver2016jupyter,
virtanen2020scipy} 
and multiple Zarr implementations (see Methods).
Of particular interest for large-scale datasets are
Dask~\cite{rocklin2015dask} and 
Xarray~\citep{hoyer2017xarray} which provide powerful abstractions
for distributed computing and can be
straightforwardly used to perform interactive analysis
of VCF Zarr datasets,
with computations transparently distributed over a cluster.
As shown in our benchmarks, lower-level computations
in which numerical operations are applied on a chunk-by-chunk
basis are also straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers can 
choose the appropriate level of abstraction and tool chain
on a case-by-case basis.

[NOTE: still trying to figure out how to string this together. 
Want to finish on a positive, not too technical note.]

and is gaining increasing
popularity in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Zarr has the potential to act as a unified, cloud-native 
efficient storage layer, freeing the next generation of 
bioinformatics tools
from the traditional burdens of text parsing and format
conversion.
